\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2016-11-18 submission/Manuscript.Rmd

Kaitlin J. Flynn^1^, Nicholas Lesniak^1^, Alyxandria M. Schubert^2^, Hamide Sinani^?^, and Patrick D. Schloss^1$\dagger$^

@@

@@

**also want something in the intro about severity maybe, and asymptomatic colonization. or in discussion?**

@@

**Microbes present in the gut prior to infection are predictive of _C. difficile_ colonization levels** Previous work in our group has demonstrated that the microbes present in the mouse microbiome prior to _C. difficile_ challenge can predict future colonization levels (cite Alyx). To determine if the human-associated microbiome in the mouse could similarly predict _C. difficile_ colonization we employed a Random Forest machine learning algorithm similar to the one used in previous studies. Relative abundance of OTUs with greater than 1% abundance at day 0 were used as input to the model. We found that the model explained X% of the variance in colonization levels of _C. difficile_ (Fig 3A). By refining the model to include only the top X predictive OTUs, the model explained X% of the variance. Partial dependency plots of the top 6 OTUs predictive of _C. difficile_ colonization reveal that colonization is positively associated with _Bacteriodetes_, _Parabacteriodes_ and _Porphyromonadaceae_ (really?) (Fig 3B). Colonization was negatively associated with the presence of _Lachnospiraceae_ (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C. difficile_ (Fig 2A, 3B). These results confirm that as in mouse microbiome studies, the human-associated gut community prior to challenge is predictive of _C. difficile_ colonization levels.

@@

Here we were able to colonize germ-free mice with a variety of human stool communities from diarrheal or healthy people. After allowing these communities to colonize the mice for 2 weeks, they were challenged with _C. difficile_. It is important to note that the mouse gut environment is different from that of the human, and as such exerts a selective pressure on the human stool microbial community that stabilizes over the 2 weeks. Thus these mice cannot be considered "humanized" but rather containing human-associated microbes. Interestingly, clinical status was not associated with proclivity to severe _C. difficile_ infection as some healthy donor mice were susceptible. A surprising finding was that none of the human-associated mouse communities were completely resistant to colonization by _C. difficile_. This is in contrast to numerous reports of mouse microbiome studies (cite) and the knowledge that in general healthy humans without gut perturbations are resistant to _C. difficile_ infection (cite). On the other hand, it is known that humans, especially health care employees, can be asymptomatic carriers of _C. difficile_ (though at what CFU?!). This may explain the fact that the majority of our groups of mice were colonized by _C. difficile_ but did not display outward signs of illness or infection-associated weight loss (Fig 2).

•	No donors were colonization resistant, discuss donor differences 
severity, asymptomatic colonzation discussion? 
•	Discuss prediction methods and outcomes - Nick
•	Discuss potential mechanisms for interesting OTUs - Nick
•	Discuss different strain results
•	Future work blah blah

@@

**DNA isolation and sequencing** Bacterial DNA was isolated from banked donor stool and each mouse stool pellet using the MO-BIO PowerSoil DNA isolation kit. This DNA was used as template for

•	Bacteria/plating 
•	Sequencing
•	Data analysis, code availability
•	Machine learning models

##Acknowledgments## 
Lab, sequencing core, Jhansi

>>>>

Kaitlin J. Flynn^1+^, Nicholas Lesniak^1+^, Alyxandria M. Schubert^2^, Hamide Sinani^?^, and Patrick D. Schloss^1$\dagger$^

@@

\+ These authors contributed equally to this work

@@

@@

**Nick will be updating the next two sections**
**Microbes present in the gut prior to infection are predictive of _C. difficile_ colonization levels** Previous work in our group has demonstrated that the microbes present in the mouse microbiome prior to _C. difficile_ challenge can predict future colonization levels (cite Alyx). To determine if the human-associated microbiome in the mouse could similarly predict _C. difficile_ colonization we employed a Random Forest machine learning algorithm similar to the one used in previous studies. Relative abundance of OTUs with greater than 1% abundance at day 0 were used as input to the model. We found that the model explained X% of the variance in colonization levels of _C. difficile_ (Fig 3A). By refining the model to include only the top X predictive OTUs, the model explained X% of the variance. Partial dependency plots of the top 6 OTUs predictive of _C. difficile_ colonization reveal that colonization is positively associated with _Bacteriodetes_, _Parabacteriodes_ and _Porphyromonadaceae_ (really?) (Fig 3B). Colonization was negatively associated with the presence of _Lachnospiraceae_ (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C. difficile_ (Fig 2A, 3B). These results confirm that as in mouse microbiome studies, the human-associated gut community prior to challenge is predictive of _C. difficile_ colonization levels.

@@

Here we were able to colonize germ-free mice with a variety of human stool communities from diarrheal or healthy people. After allowing these communities to colonize the mice for 2 weeks, they were challenged with _C. difficile_. It is important to note that the mouse gut environment is different from that of the human, and as such exerts a selective pressure on the human stool microbial community that stabilizes over the 2 weeks. Thus these mice cannot be considered "humanized" but rather containing human-associated microbes. Interestingly, clinical status was not associated with proclivity to severe _C. difficile_ infection as some healthy donor mice were susceptible. A surprising finding was that none of the human-associated mouse communities were completely resistant to colonization by _C. difficile_. This is in contrast to numerous reports of mouse microbiome studies (cite) and the knowledge that in general healthy humans without gut perturbations are resistant to _C. difficile_ infection (cite). On the other hand, it is known that humans, especially health care employees, can be asymptomatic carriers of _C. difficile_ (Furyura-Kanamari). This may explain the fact that the majority of our groups of mice were colonized by _C. difficile_ but did not display outward signs of illness or infection-associated weight loss (Fig 2).

•	Discuss prediction methods and outcomes - Nick - we have bugs that predict severity, but do we have bugs that predict mild/asymptomatic? probably, could discuss here
•	Discuss potential mechanisms for interesting OTUs - Nick

Finally, our pilot experiments testing the efficiency of infection with different _C. difficile_ strains have allowed for testing community versus pathogen factors of infection. Our results are consistent with previous reports that the composition of the gut microbiome determines if an individual is successfully infected or not (cite Alyx and others). _C. difficile_ strains are known to vary in pathogenecity, primarily via the amount of toxin that is produced during infection (cite). In these pilot experiments, we challenged a susceptible community with another pathogenic strain of _C. difficile_ to confirm that the susceptibility lied in the community and not in the specific challenge strain. In contrast, we exposed a resistant/asymptomatic community to a relatively weak strain of _C. difficile_ and found that the community could be colonized by this one as well-- but not cause disease. This suggests that this colony is prone to colonization by any _C. difficile_ strain. Further experiments will test if infecting susceptible communities with weak strains allows the mice to survive. It has been recently reported that carriage of a non-toxigenic strain of _C. difficile_ can prevent additional colonization by a pathogenic strain, suggesting some role of niche-filling or immunity this provides (cite cite). It would be interesting to see if we could make a susceptible community resistant to infection and disease by a pathogenic _C. difficle_ strain in our human-associated microbiome mice. Further work with human-associated communities in mice and mathematical modeling will allow for uncovering of complex human-associated disease processes and dynamics as they relate to human disease.

@@

**DNA isolation and sequencing** Bacterial DNA was isolated from banked donor stool and each mouse stool pellet using the MO-BIO PowerSoil DNA isolation kit. This DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (cite Jim paper).

•	Sequence curation/mothur, data analysis, code + availability of data(SRA) and code (Github) - Nick

##Acknowledgments## 
We would like to thank members of the Schloss lab and Jhansi Leslie for analysis advice and the Microbial Systems Laboratory at the University of Michigan for sequencing assistance.

### 2016-11-18 submission/Manuscript.tex

Kaitlin J. Flynn\textsuperscript{1}, Nicholas
Lesniak\textsuperscript{1}, Alyxandria M. Schubert\textsuperscript{2},
Hamide Sinani\textsuperscript{?}, and Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

@@

@@

@@

\textbf{The microbiome predicts \emph{C. difficile} severity.}

\textbf{Results to be written}

\begin{enumerate}
\def\labelenumi{\alph{enumi}.}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Figure 4: Random Forest predicts CDI severity, predictive OTUs
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Propensity for severe CDI is community-dependent and
  strain-independent
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\alph{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Figure 5: Infection of mice with different C. difficile strains.
\end{enumerate}

@@

• Restate results, • caveats about mouse weights • No donors were
colonization resistant, discuss donor differences • Discuss prediction
methods and outcomes • Discuss potential mechanisms for interesting OTUs
• Discuss different strain results • Future work blah blah

@@

\textbf{Human samples} Donor stool samples were chosen from a sample
bank collected from patients from October 2010 to November 2012 at the
University of Michigan. Sample collection methods and study was approved
by the Institutional Review Board at the University of Michigan. Stool
samples from University of Michigan Health System patients were
submitted to the microbiology lab for \emph{C. difficile} testing.
\emph{C. difficile} presence was measured using WHATEVER- Cdiff quik
check? toxin AB?. Samples from these patients were stored in Cary-Blair
medium at -80C. All patients consented to participation in the study.
Donor stool for this study was selected from one patient that tested
positive for \emph{C. difficile} infection and four patients that
presented with diarrhea that tested negative for \emph{C. difficile}.
Healthy donor stool was collected from volunteer donors living in the
Ann Arbor, MI area.

>>>>

Kaitlin J. Flynn\textsuperscript{1+}, Nicholas
Lesniak\textsuperscript{1+}, Alyxandria M. Schubert\textsuperscript{2},
Hamide Sinani\textsuperscript{?}, and Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

@@

+ These authors contributed equally to this work

@@

\textbf{also want something in the intro about severity maybe, and
asymptomatic colonization. or in discussion?}

@@

\textbf{The microbiome predicts \emph{C. difficile} infection severity.}
As \emph{C. difficile} colonization levels can be predicted by the
microbiome composition on day 0, we postulated that severity of disease
could also be predicted from day 0. Using weight loss as a proxy for
severity, we built a Random Forest classification model based on
relative abundance to determine which community members were associated
with either mild or severe disease. We optimized the model using the
AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with
maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously
in our group to classify colon cancer outcomes (cite Niel 2015 gen med).
The optimized model used X number of OTUs and identified several that
predicted increased weightloss and subsequent disease severity (Fig 4A,
4B). The populations that were most commonly found in mice that
succumbed to \emph{C. difficile} infection were members of the
\emph{Clostridiales}, \emph{Bacilliales}, \emph{Streptococcus}
(Lactobacilles?) and \emph{Bacteriodes}. (need more OTUs here, or number
that were used for model and then group by rank). The AUC was calculated
using the Out-Of-Bag (OOB) error for each sample (Fig 4A).
Cross-validation using 10-fold, leave-one-mouse-out and
leave-one-cage-out resulted in AUCs not significantly different from the
OOB (Fig 4A). This model reveals that the levels of X OTUs on day 0 can
predict disease outcome in mice with human-associated microbiomes.
(probably also going to want a supplemental table for predictive OTUs
for this and Fig 3)

\textbf{Susceptibility to infection is community-dependent.} \emph{C.
difficile} strains are known to vary in disease phenotypes based both on
pathogenicity factors of the strain and host microbiome composition
(cite someone). To test the susceptibility of human-derived microbiomes
to \emph{C. difficile} infection, two donor-mouse microbiomes that were
previously shown to be resistant to \emph{C. difficile} strain 431
induced weight loss (Fig 2B) (DA00396, DA00430) and one that was
suscpetible to severe disease (DA00578) were independently challenged
with one of three unique \emph{C. difficile} strains (table or reference
for toxin prod, etc). As done previously, \emph{C. difficile} stool CFU
was enumerated daily and mouse weight-loss tracked. Infection of mice
with resistant communities with virulent strain 458 or the weaker strain
299 resulted in colonization to levels near that of 431 (Fig 5A) and
these mice did not experience severe weight loss or show clinical signs
of disease (Fig 5B). In contrast, infection of mice harboring
microbiomes susceptible to severe disease (DA00578, Fig 1B) with strain
395 resulted in colonization and severe weight loss by two days
post-infection (Fig 5A, 5B). Strain 395 is similarly virulent as strain
431, and shows a similar infection phenotype in these mice. These
preliminary results suggest that human-associated microbiomes may be
more predictive of disease severity than the \emph{C. difficile} strain.

@@

Study of human-associated microbiomes and their interactions with
\emph{C. difficile} has been limited by the lack of an appropriate model
and available samples. Here we colonized 16 groups of mice with human
donor stool from patients with different clinical profiles or healthy
volunteers. This created a diverse set of human-associated microbiomes
in the mice that allowed for testing \emph{C. difficile} infection
dynamics and community interactions. Our results showed that both
healthy and diarrheal-associated human communities were susceptible to
severe disease. Additionally, our results confirmed our and others
previous studies that proved that resistance to \emph{C. difficile}
disease is dependent on the entire microbial community and not just one
or two species (cite alyx). Further, we were able to learn
classification models that identified specific OTUs within susceptible
communities that predispose mice to infection. By subsequently infecting
susceptible or resistant communities with varying strains of \emph{C.
difficile} we confirmed that the initial state of the human-associated
microbiome is more predictive of disease severity than the strain used.

Here we were able to colonize germ-free mice with a variety of human
stool communities from diarrheal or healthy people. After allowing these
communities to colonize the mice for 2 weeks, they were challenged with
\emph{C. difficile}. It is important to note that the mouse gut
environment is different from that of the human, and as such exerts a
selective pressure on the human stool microbial community that
stabilizes over the 2 weeks. Thus these mice cannot be considered
``humanized'' but rather containing human-associated microbes.
Interestingly, clinical status was not associated with proclivity to
severe \emph{C. difficile} infection as some healthy donor mice were
susceptible. A surprising finding was that none of the human-associated
mouse communities were completely resistant to colonization by \emph{C.
difficile}. This is in contrast to numerous reports of mouse microbiome
studies (cite) and the knowledge that in general healthy humans without
gut perturbations are resistant to \emph{C. difficile} infection (cite).
On the other hand, it is known that humans, especially health care
employees, can be asymptomatic carriers of \emph{C. difficile}
(Furyura-Kanamari). This may explain the fact that the majority of our
groups of mice were colonized by \emph{C. difficile} but did not display
outward signs of illness or infection-associated weight loss (Fig 2).

• Discuss prediction methods and outcomes - Nick - we have bugs that
predict severity, but do we have bugs that predict mild/asymptomatic?
probably, could discuss here • Discuss potential mechanisms for
interesting OTUs - Nick

Finally, our pilot experiments testing the efficiency of infection with
different \emph{C. difficile} strains

Discuss different strain results

• Future work/conclusion section/sentence

@@

\textbf{Human samples} Donor stool samples were chosen from a sample
bank collected from patients from October 2010 to November 2012 at the
University of Michigan. Sample collection methods and study was approved
by the Institutional Review Board at the University of Michigan. Stool
samples from University of Michigan Health System patients were
submitted to the microbiology lab for \emph{C. difficile} testing.
\emph{C. difficile} presence was measured using WHATEVER- Cdiff quik
check? toxin AB?. Samples from these patients were stored in Cary-Blair
medium at -80C. All patients consented to participation in the study.
Donor stool for this study was selected from one patient that tested
positive for \emph{C. difficile} infection and five patients that
presented with diarrhea and tested negative for \emph{C. difficile}.
Healthy donor stool was collected from adult volunteers living in the
Ann Arbor, MI area. 100ul of stool slurry was gavaged into each mouse.

\textbf{\emph{Clostridium difficile} and sample processing} For all
experiments, 100 spores of \emph{Clostridium difficile} were used to
orally infect mice as described previously (cite alyx). The \emph{C.
difficile} strains used were acquired from spore stocks of (WHERE) and
comprised of strains 431, 395, 299 and 468 (cite?). After gavage, the
remaining inoculum was diluted and plated to confirm infection dose.
Bacteria were cultured anaerobically on TCCFA plates and incubated at
37C. During infection, two fresh stool pellets were collected from each
mouse daily. One pellet was resuspended in anaerobic phosphate-buffered
saline and diluted and plated on TCCFA plates to enumerate \emph{C.
difficile} CFU. The other pellet was frozen at -80C until the end of the
experiment.

\textbf{DNA isolation and sequencing} Bacterial DNA was isolated from
banked donor stool and each mouse stool pellet using the MO-BIO
PowerSoil DNA isolation kit. This DNA was used as template for
